---
title: "Introduction to Object-Oriented Programming and S3 System in R"
author: "S.Ravichandran ( https://github.com/ravichas/OOP-S3-in-R )"
date: "April 27, 2019"
output: pdf_document 
theme: flatly
---

# based on Carl Broman's code 
<style type="text/css">

ul,ol {
  margin: 0 0 0 35px;
}

body, td {
   font-size: 14px;
}
code.r{
  font-size: 20px;
}
pre {
  font-size: 20px
}
</style>



```{r setup, include=FALSE}
# set global chunk options: images will be 7x5 inches
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
options(digits = 4)
knitr::opts_chunk$set(echo = TRUE)
```


### Preliminary information about object types in R

Let us create a logical object, x.
```{r logicaltype}
(x <- TRUE) # logical
print(class(x))
```

Let us create a list, also called x. 
```{r numericaltype}
(x <- list(nums = 1:10, 
          chars = c("one","two","three"),
          ints = c(1L,2L,3L)
          ))
print(class(x))
```

BMI is a data.frame with four variables, Gender, Height, Weight and Age.
```{r dftype}
(BMI <- 	data.frame(
   Gender = c("Male", "Male","Female"), 
   Height = c(153.1, 173.6, 165.0), 
   Weight = c(81,93, 78),
      Age = c(42,38,26)
))
print(class(BMI))
```

### Hands-on 1

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# needed libraries
packages <- c("pryr", "sloop","magrittr","dplyr","ggplot2")
if ( length(missing_pkgs <- setdiff(packages, rownames(installed.packages()))) > 0) {
  message("Installing missing package(s): ", paste(missing_pkgs, collapse = ", "))
  install.packages(missing_pkgs)
}

library(magrittr)
library(dplyr)
library(ggplot2)
library(pryr)
library(sloop)
```

One of the important concept of OOP is functions can 
respond in different ways depending on the input object type.
To explain this concept, let us create the following objects:

  * Numeric vector of 10 random numbers
  * Categorical vector of length 6 
  * A linear model object
  
First, let us create a numerical vector with 10 elements.
```{r numeric-var, eval=TRUE }
set.seed(111)
(x_num <- rnorm(10) )
```


Next, we build a categorical vector with 6 elements.
```{r factor-var, eval=TRUE}
(x_fac <- factor(c("A", "B", "A", "C", "A", "B")))
```


Finally, let us create a linear model variable. 
But, first let us create two variables x and y
```{r model-vars, eval=TRUE}
# setting seed
set.seed(123)
(x <- 1:10)
(y <- jitter(x, amount = 2))
```

```{r var-plot, eval=TRUE}
data.frame(x, y) %>% ggplot(aes(x, y)) + 
  geom_point() + 
  geom_smooth(method = "lm", col = "red", se = FALSE)
```

Build a model
```{r model-object, eval=TRUE}
model <- lm(y ~ x)
model
```


### Behavior of summary function on different class of objects

```{r summary-numeric}
x_num
summary(x_num)
```

```{r summary-factor}
x_fac
summary(x_fac)
```

```{r summary-model}
model
summary(model)

```

### How does R distinguish types of variables? 

what command(s) can be used for this task?

```{r}
# matrix 
(int_mat <- matrix(1:12, nrow = 4, ncol = 3 )) # column major

# determine the variable
class(int_mat) # obj is a matrix

# what type of matrix (elements are of what type)
typeof(int_mat) # int matrix; content of the matrix
```

```{r}
(float_mat <- matrix(rnorm(12), nrow = 4, ncol = 3))
class(float_mat) # matrix
typeof(float_mat) # double; type of var that makes up matrix
# c code; in C floating point #s are double
```


### Hands-on 2: 

Interrogation of objects to see whether they are S3 objects

```{r}
(int_mat <- matrix(1:12, nrow = 4, ncol = 3 )) # column major
sloop::otype(int_mat) # package::command(object)

head(mtcars)
sloop::otype(mtcars)

```

### S3 & R6: How to assign classes?

* Can I override the class? 
   * Yes
* And as expected, it wont break the functionality
* Can I woverride the type? 
   * No 

```{r}
x_num
class(x_num)
typeof(x_num)

(class(x_num) <- "random-numbers")

# the class that we have added has become an attribute
x_num
# we cannot override typeof 
typeof(x_num)
is.numeric(x_num) # no matter what the class says
```

### S3 & R6: Function overloading

S3 exists so that we dont have to write many many functions to take
care of different data types.

How does it work?

* S3 splits a function into generic and method functions.
* Methods named generic.class (Ex. print.Date)

Example of generic functions are print, summary etc.

```{r}

(x_Date <- Sys.Date()) # "YYYY-MM-DD"
class(x_Date) # "Date"

print(x_Date) # "YYYY-MM-DD", 2019-03-26
# is same as calling print.Date
print.Date(x_Date) 

# Let us explore the print function
print

```

print function is just a simple one line function.
You can ignore the last two lines that shows the memory 
location and the object  environment. 
print function calls UseMethod("print") to provide the final output.


### What methods exist for a generic function?

* For example, for the generic function what methods are available
* generic.class1, generic.class2, generic.class3

Exmaple. print (generic), print.data.frame, print.Date etc.
```{r What-methods-for-generic}
head(methods(print)) # too many methods 
```

### What methods are available for a given class of an object? 

* The methods could be coming from different generic classes. 
  For example, generic1.class, generic2.class etc.
* Note this methods call for this case will return both S3 and s4 objects. 

```{r}
# gives both S3 and S4
methods(class = "lm") # or methods(class=lm)
# of them which ones are S3
.S3methods(class = "lm")
```

### Is the object/function generic or method? 

```{r}
pryr::is_s3_generic("print") # TRUE
pryr::is_s3_method("print") # FALSE
pryr::is_s3_method("print.Date") # TRUE
```
Let us define our object.

```{r}
(people <- c("Frank Blanchard",
            "Andrea Gnuschke",
            "Max Cole",
            "Maryellen Hackett",
            "Victoria Brun",
            "Jonathan Summers",
            "Christopher Worthington",
            "Samuel Lopez",
            "Richard Frederickson",
            "Chris Hu") )
class(people)
(class(people) <- "InsiteGroup")
```

Suppose, we want to write an S3 function that gets the 
first name from the InsiteGroup object.

```{r}
GetFirst <- function(obj) { 
  UseMethod("GetFirst",obj) 
  }

# create methods function 
GetFirst.InsiteGroup <- function(obj) { 
  return(obj[1])
}

# create default function 
GetFirst.default <- function(obj){
  cat("This is a generic class\n")
  # do something
  }
methods(class = "InsiteGroup")

GetFirst(people)

```

If no suitable methods can be found for a generic, 
then an error is thrown. For example, at the moment, get_n_elements() only has 2 methods available. If you 
pass a data.frame/matrix to get_n_elements() instead, 
you'll see an error. One could use generic.default to 
deal with all the missing class of objects.

### Can variables have more than one class? 

```{r S3-MultipleClasses}
(human <- "laugh")
# less specific to more specific; final default class,character
class(human) <- c("mammalia","eukaryota","character")

# create a generic method for who_am_i
who_am_i <- function(x, ...) {
  UseMethod("who_am_i")
}

# create mammalia method for who_am_i
who_am_i.mammalia <- function(x, ...) {
  # let us write a message 
  message("I am a Mammal")
  NextMethod("x")
}

# create eukarota method for who_am_i
who_am_i.eukaryota <- function(x, ...) {
  # let us write a message 
  message("I am a Eukaryote")
  NextMethod("x")
}

# finally one for character method
who_am_i.character <- function(x, ...) {
  # let us write a message 
  message("I am a simple character!")
  # since this is the last, no NextMethod
}


# call human to see all the 3 messages are displayed
class(human)

who_am_i(human) 

```





 